---
title: "STAT545_hw02_exploration"
output: github_document
---

```{r}
library(gapminder)
library(tidyverse)
```
# Smell test the data

What is the gapminder object?
```{r}
typeof(gapminder)
```

What is its class?

```{r}
class(gapminder)
```

How many variables/columns?

```{r}
ncol(gapminder)
```

How many rows/observations?

```{r}
nrow(gapminder)
```

Can you get these facts about “extent” or “size” in more than one way? Can you imagine different functions being useful in different contexts?

```{r}
dim(gapminder)
```

nrow and ncol functions could be useful if creating loops that require the number of rows or columns. dim function could be useful for matrix calculations

What data type is each variable?

```{r}
sapply(gapminder, class)
```

# Explore individual variables

What are possible values (or range, whichever is appropriate) of each variable?

```{r}
range(gapminder$year)
```

```{r}
range(gapminder$lifeExp)
```

```{r}
table(gapminder$continent)
```


What values are typical? What’s the spread? What’s the distribution? Etc., tailored to the variable at hand.

```{r}
summary(gapminder$year)
```

```{r}
summary(gapminder$lifeExp)
```

```{r}
summary(gapminder$continent)
```

Feel free to use summary stats, tables, figures. We’re NOT expecting high production value (yet).

# Explore various plot types

Make a few plots, probably of the same variable you chose to characterize numerically. You can use the plot types we went over in class (cm006) to get an idea of what you’d like to make. Try to explore more than one plot type. Just as an example of what I mean:

A scatterplot of two quantitative variables.

```{r}
filter(gapminder, country == "Afghanistan" | country == "Canada") %>% 
ggplot(aes(year,lifeExp)) +
  geom_point(aes(colour = country), size = 3, shape = 15) +
  geom_smooth(method = "lm", se = FALSE, aes(colour = country)) +
  theme_classic()
```


A plot of one quantitative variable. Maybe a histogram or densityplot or frequency polygon.

```{r}
filter(gapminder, continent == "Africa" & year == 2007) %>% 
  ggplot(aes(gdpPercap)) +
  geom_histogram(bins = 50, colour = "black", fill = "blue", alpha = .5) +
  theme_classic()
```



A plot of one quantitative variable and one categorical. Maybe boxplots for several continents or countries.


```{r}

filter(gapminder, continent == "Americas") %>% 
  ggplot(aes(country, gdpPercap)) +
  geom_boxplot() +
  theme(axis.text.x  = element_text(angle=90, vjust=1, size=7))

```


# But I want to do more\!

#An analyst intended to filter *gapminder* data into the entries for the two countries, Afghanistan and Rwanda, by using the following code chunk:

```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```

#By using the c() function, R combined the entries into a vector of the same size as one country's entries. In this case, the function seems to have split them along the year columnm, XXX2 dates from Rwanda and XXX7 dates from Afghanistan, so half of the entries being from each.

#One way to achieve the intended result would have been by filtering for Rwanda entries and filtering for Afghanistan entries.

```{r}
filter(gapminder, country == "Rwanda" | country == "Afghanistan")
```

#Here are some additional *dplyr* functions and use of *knitr* to make a simple table.

```{r}
filter(gapminder, year == 2007) %>% 
  arrange(desc(gdpPercap)) %>% 
  select(country, lifeExp, pop, gdpPercap) %>% 
  knitr::kable(digits = 0,caption = "Table 1: Countries ranked by GDP per capita [2007].")
```

